Recent Advances in Networking

### 2013-07-26 Compiled/chapter9/Content/flood_rtg.tex
DOIs updated in chapt 9

Together with 1-hop neighbors, additional information may be included in Hello messages -- in particular, the cost of the links towards the listed neighbors, when metrics other than hop-count (see section \ref{ss:metrics}) is used. Exchange of Hello messages is typically done periodically, although some events may trigger non-periodic Hellos ({\em e.g.}, changes in the topology). \ \\ \ \\ %({\em e.g.} ETX \cite{etx}, based on the number of expected transmissions to send successfully a packet over a link, or EDR \cite{edr}, that estimates the data rate at which packets can be sent) is used.
%
The {\em Mobile Ad Hoc Network Neighborhood Discovery Protocol} (NHDP, RFC 6130) \cite{NHDP-RFC6130} is the main ND protocol for spontaneous wireless networks. It is used as auxiliary protocol by other routing protocols that need neighborhood information to take their decisions, such as OLSRv2 (see section \ref{sec:olsr}). In NHDP, Hello messages are exchanged periodically and they contain the id of the originating router and the list of its 1-hop neighbors. The IETF also standardized an optimization of NDP for IPv6 over Low-Power Wireless Personal Area Networks (6LoWPANs). This optimization, specified in RFC 6775, adapts the operation of NDP to the lossy conditions of communications and the low-power device constraints of LoWPANs. This is done, for instance, by avoiding unsollicited messages (such as periodic router announcements), reducing the use of multicast for address resolution, limiting the duplicate address detection checks, or enabling better compression algorithms \cite{rfc6775}. \ \\ \ \\
%
%lossy, low-power, low-bit-rate, short-range; with many nodes saving energy with long sleep periods. Multicast as used in IPv6 Neighbor Discovery (ND) [RFC4861] is not desirable in such a wireless low-power and lossy network. Moreover, LoWPAN links are asymmetric and non-transitive in nature. 
%
Other routing protocols include their own Hello mechanism. This is the case of AODV \cite{AODV-RFC3561} (see section \ref{ss:aodv}) or OSPF and its MANET extensions \cite{rfc5449,rfc5614,rfc5820} (see section \ref{ss:ospfmanet}). In the last case, some approches have been explored in order to avoid redundant notifications and hence reduce control traffic by only reporting changes in the neighborhood occurred since the last Hello transmission: this principle leads to the {\em incremental Hellos} mechanism used in the Overlapping Relays extension of OSPF (OR/SP \cite{rfc5820}) and the {\em differential Hellos} mechanism used in the MANET Designated Routers extension of OSPF (OSPF-MDR \cite{rfc5614}). However, experiments show that the potential benefits (mostly, saved amount of traffic) of these two mechanisms are not significant, in particular when compared with the additional complexity they introduce in the corresponding protocols \cite{aircc}.

@@

The fact that flooded messages are forwarded simultaneously by wireless interfaces receiving them through the same wireless shared medium may cause packet collisions during the flooding procedure (depending on the network topology at the time of flooding). Unlike other packet transmissions for which the collision probability may vary depending on the traffic pattern, these flooding collisions are {\em systematic} and will occur, for a given network topology and flooding algorithm, any time that the source node floods a new message (in the example, any time $A$ floods a message). \ \\ \ \\
%
This effect could be alleviated by allowing routers to wait a random amount of time (denominated {\bf jitter}) before retrasmitting a flooded message, in order to reduce the probability of concurrent transmissions by neighboring wireless interfaces. This technique is known as {\bf jittering}, and has been standardized by the IETF in RFC 5148 \cite{jitter-RFC5148}. The recommendation from RFC 5148 is that delays are selected following an uniform distribution between $0$ and a maximum jitter value, $MAX\_JITTER$. Figure \ref{f:jitter} illustrates the effect of jittering techniques in the network example of Figure \ref{f:syspcol}. In the example, node \textit{A} is flooding a packet to all the other nodes. When node \textit{B} and \textit{C} receive the packet from \textit{A}, instead of retransmitting the packet immediately, they wait a random delay. In this way, simultaneous transmission of $B$ and $C$ (which can cause collision at $D$ in this case) can be avoided. \ \\ \ \\
%
\begin{figure}[htb]
\centering
\includegraphics[width=0.6\textwidth]{Figures/jitter4flooding-crop.pdf}
\caption{Use of jitter for flooding. Node \textit{A} is flooding a packet in a network. Node \textit{B} and \textit{C} wait a random delay before the packet is retransmitted. The dashed overlapping arrows represent the packet collision that would occur if no jitter were used.}
\label{f:jitter}
\end{figure}
%
Although jittering can be theorically implemented at different layers of the protocol stack, it has been shown that its use in layers upper than L3 brings little benefit \cite{jitter-broadcast}. As the problem of systematic collisions affects every L3 routing protocol using wireless flooding, jittering techniques can be implemented by different protocols. The addition of random delays in flooded packets impacts differently in proactive and reactive protocols, given the different use of flooding in both routing strategies. Other jittering effects are due to the specifities of the techniques employed in each case. In proactive protocols where jitter is used (as described in RFC 5148 \cite{jitter-RFC5148}, which recommends to introduce random delays and piggyback all pending messages when a transmission is scheduled), such as OLSR or the OSPF extensions for MANETs, jittering leads to longer LSA messages -- this may cause additional packet collisions, if jitter values are not configured properly \cite{INFOCOM-jitter,jitter-wpc}. In reactive protocols such as AODV, where jitter is used for Route Request (RREQ) flooding, the addition of random delays may lead to suboptimal path selections, which can be minimized by adapting the random distribution used for determining jitter values \cite{Infocom12, wiopt}.

>>>>

Together with 1-hop neighbors, additional information may be included in Hello messages -- in particular, the cost of the links towards the listed neighbors, when metrics other than hop-count (see section \ref{ss:metrics}) is used. Exchange of Hello messages is typically done periodically, although some events may trigger non-periodic Hellos ({\em e.g.}, changes in the topology). \ \\ \ \\ %({\em e.g.} ETX \cite{etx}, based on the number of expected transmissions to send successfully a packet over a link, or EDR \cite{edr}, that estimates the data rate at which packets can be sent) is used.
%
The {\em Mobile Ad Hoc Network Neighborhood Discovery Protocol} (NHDP, RFC 6130) \cite{rfc6130} is the main ND protocol for spontaneous wireless networks. It is used as auxiliary protocol by other routing protocols that need neighborhood information to take their decisions, such as OLSRv2 (see section \ref{sec:olsr}). In NHDP, Hello messages are exchanged periodically and they contain the id of the originating router and the list of its 1-hop neighbors. The IETF also standardized an optimization of NDP for IPv6 over Low-Power Wireless Personal Area Networks (6LoWPANs). This optimization, specified in RFC 6775, adapts the operation of NDP to the lossy conditions of communications and the low-power device constraints of LoWPANs. This is done, for instance, by avoiding unsollicited messages (such as periodic router announcements), reducing the use of multicast for address resolution, limiting the duplicate address detection checks, or enabling better compression algorithms \cite{rfc6775}. \ \\ \ \\
%
%lossy, low-power, low-bit-rate, short-range; with many nodes saving energy with long sleep periods. Multicast as used in IPv6 Neighbor Discovery (ND) [RFC4861] is not desirable in such a wireless low-power and lossy network. Moreover, LoWPAN links are asymmetric and non-transitive in nature. 
%
Other routing protocols include their own Hello mechanism. This is the case of AODV \cite{rfc3561} (see section \ref{ss:aodv}) or OSPF and its MANET extensions \cite{rfc5449,rfc5614,rfc5820} (see section \ref{ss:ospfmanet}). In the last case, some approches have been explored in order to avoid redundant notifications and hence reduce control traffic by only reporting changes in the neighborhood occurred since the last Hello transmission: this principle leads to the {\em incremental Hellos} mechanism used in the Overlapping Relays extension of OSPF (OR/SP \cite{rfc5820}) and the {\em differential Hellos} mechanism used in the MANET Designated Routers extension of OSPF (OSPF-MDR \cite{rfc5614}). However, experiments show that the potential benefits (mostly, saved amount of traffic) of these two mechanisms are not significant, in particular when compared with the additional complexity they introduce in the corresponding protocols \cite{aircc}.

@@

The fact that flooded messages are forwarded simultaneously by wireless interfaces receiving them through the same wireless shared medium may cause packet collisions during the flooding procedure (depending on the network topology at the time of flooding). Unlike other packet transmissions for which the collision probability may vary depending on the traffic pattern, these flooding collisions are {\em systematic} and will occur, for a given network topology and flooding algorithm, any time that the source node floods a new message (in the example, any time $A$ floods a message). \ \\ \ \\
%
This effect could be alleviated by allowing routers to wait a random amount of time (denominated {\bf jitter}) before retrasmitting a flooded message, in order to reduce the probability of concurrent transmissions by neighboring wireless interfaces. This technique is known as {\bf jittering}, and has been standardized by the IETF in RFC 5148 \cite{rfc5148}. The recommendation from RFC 5148 is that delays are selected following an uniform distribution between $0$ and a maximum jitter value, $MAX\_JITTER$. Figure \ref{f:jitter} illustrates the effect of jittering techniques in the network example of Figure \ref{f:syspcol}. In the example, node \textit{A} is flooding a packet to all the other nodes. When node \textit{B} and \textit{C} receive the packet from \textit{A}, instead of retransmitting the packet immediately, they wait a random delay. In this way, simultaneous transmission of $B$ and $C$ (which can cause collision at $D$ in this case) can be avoided. \ \\ \ \\
%
\begin{figure}[htb]
\centering
\includegraphics[width=0.6\textwidth]{Figures/jitter4flooding-crop.pdf}
\caption{Use of jitter for flooding. Node \textit{A} is flooding a packet in a network. Node \textit{B} and \textit{C} wait a random delay before the packet is retransmitted. The dashed overlapping arrows represent the packet collision that would occur if no jitter were used.}
\label{f:jitter}
\end{figure}
%
Although jittering can be theorically implemented at different layers of the protocol stack, it has been shown that its use in layers upper than L3 brings little benefit \cite{jitter-broadcast}. As the problem of systematic collisions affects every L3 routing protocol using wireless flooding, jittering techniques can be implemented by different protocols. The addition of random delays in flooded packets impacts differently in proactive and reactive protocols, given the different use of flooding in both routing strategies. Other jittering effects are due to the specifities of the techniques employed in each case. In proactive protocols where jitter is used (as described in RFC 5148 \cite{rfc5148}, which recommends to introduce random delays and piggyback all pending messages when a transmission is scheduled), such as OLSR or the OSPF extensions for MANETs, jittering leads to longer LSA messages -- this may cause additional packet collisions, if jitter values are not configured properly \cite{INFOCOM-jitter,jitter-wpc}. In reactive protocols such as AODV, where jitter is used for Route Request (RREQ) flooding, the addition of random delays may lead to suboptimal path selections, which can be minimized by adapting the random distribution used for determining jitter values \cite{Infocom12, wiopt}.
